---
title: 'DATS6101 Project 1'
author: 'Group 6'
date: 'r Sys.Date()'
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=FALSE}
#Import required libraries
library(ezids)
library(ggplot2)
library(dplyr)

#Setup knitter options
knitr::opts_chunk$set(warning = F, message = F)
options(scientific=T, digits = 3)
```

# Introduction
## About the Project
**In the hotel industry, effective management of bookings is crucial for optimizing revenue and ensuring operational efficiency. Hotels experience a variety of booking patterns, some of which include cancellations or no-shows. While some customers may cancel due to changing plans, others may not show up despite a confirmed reservation. To mitigate financial losses and maximize occupancy, it’s important for hotels to understand the underlying factors that contribute to these cancellations.**\
**In this project, we aim to study the factors influencing hotel booking cancellations and no-shows. By examining customer attributes such as room preferences, lead time (the number of days between the booking date and arrival date), meal plan choices, and past booking behavior, we will identify key patterns and trends. Additionally, we will analyze other variables such as special requests, parking space requirements, and market segments. Our analysis will focus on determining how these factors impact the likelihood of cancellations and assessing whether stricter policies could reduce cancellation rates.**\

## About the dataset
**For this project, we have selected a hotel booking dataset containing over 36,000 records of bookings. [INNHotelsGroup dataset](https://www.kaggle.com/datasets/mariyamalshatta/inn-hotels-group) from Kaggle This dataset offers a comprehensive overview of hotel booking patterns, making it ideal for our analysis on cancellations and no-shows.**

```{r import_dataset}
hotel_data = read.csv("DATS6101_PROJECT-1/Dataset/INNHotelsGroup.csv")
```

This dataset contains `r nrow(hotel_data)` observations of `r ncol(hotel_data)` variables. Out of these observations, `r sum(is.na(hotel_data))` rows contain null values.

## Cleaning the dataset
For our project, we are interested in majority of the variables but there are some irrelevant columns with respect to our objective and thus we will drop them.

```{r clean_data}
hotel_data[, c("no_of_adults", "no_of_children", "no_of_weekend_nights", "no_of_week_nights", 
"type_of_meal_plan", "required_car_parking_space", "room_type_reserved", "lead_time", "arrival_year", "arrival_month",  "market_segment_type", "repeated_guest", "no_of_previous_cancellations", "no_of_special_requests", "booking_status", "avg_price_per_room")]

# Remove rows with missing values
hotel_data_min = na.omit(hotel_data)
```

**After cleaning, we are left with `r nrow(hotel_data_min)` observations of `r ncol(hotel_data_min)` variables.**

```{r save_data}
write.csv(hotel_data_min,"DATS6101_PROJECT-1/Dataset/INNHotelsGroup_min.csv", row.names = F)
```