---
title: "Project1"
author: "Tycho Gormley"
date: "2024-10-06"
output: html_document
---
## Background

A significant number of hotel bookings are called off due to cancellations or no-shows. The typical reasons for cancellations include change of plans, scheduling conflicts, etc. This is often made easier by the option to do so free of charge or preferably at a low cost which is beneficial to hotel guests but it is a less desirable and possibly revenue-diminishing factor for hotels to deal with. Such losses are particularly high on last-minute cancellations.

The new technologies involving online booking channels have dramatically changed customers' booking possibilities and behavior. This adds a further dimension to the challenge of how hotels handle cancellations, which are no longer limited to traditional booking and guest characteristics.

The cancellation of bookings impact a hotel on various fronts:
1. Loss of resources (revenue) when the hotel cannot resell the room.
2. Additional costs of distribution channels by increasing commissions or paying for publicity to help sell these rooms.
3. Lowering prices last minute, so the hotel can resell a room, resulting in reducing the profit margin.
4. Human resources to make arrangements for the guests.

## Objective


## Data Description

The data contains the different attributes of customers' booking details. The detailed data dictionary is given below.


**Data Dictionary**

* Booking_ID: unique identifier of each booking
* no_of_adults: Number of adults
* no_of_children: Number of Children
* no_of_weekend_nights: Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel
* no_of_week_nights: Number of week nights (Monday to Friday) the guest stayed or booked to stay at the hotel
* type_of_meal_plan: Type of meal plan booked by the customer:
    * Not Selected – No meal plan selected
    * Meal Plan 1 – Breakfast
    * Meal Plan 2 – Half board (breakfast and one other meal)
    * Meal Plan 3 – Full board (breakfast, lunch, and dinner)
* required_car_parking_space: Does the customer require a car parking space? (0 - No, 1- Yes)
* room_type_reserved: Type of room reserved by the customer. The values are ciphered (encoded) by INN Hotels.
* lead_time: Number of days between the date of booking and the arrival date
* arrival_year: Year of arrival date
* arrival_month: Month of arrival date
* arrival_date: Date of the month
* market_segment_type: Market segment designation.
* repeated_guest: Is the customer a repeated guest? (0 - No, 1- Yes)
* no_of_previous_cancellations: Number of previous bookings that were canceled by the customer prior to the current booking
* no_of_previous_bookings_not_canceled: Number of previous bookings not canceled by the customer prior to the current booking
* avg_price_per_room: Average price per day of the reservation; prices of the rooms are dynamic. (in euros)
* no_of_special_requests: Total number of special requests made by the customer (e.g. high floor, view from the room, etc)
* booking_status: Flag indicating if the booking was canceled or not.

### Loading Data
```{r}
hotel <- read.csv("INNHotelsGroup.csv")
# copying data to another variable to avoid any changes to original data
data <- hotel
```

### View the first and last 5 rows of the dataset
```{r}
class(data)
head(data)
tail(data)
```

### Check info of the data
```{r}
str(data)
```

### Cleaning
We can see that all of the entries in the Booking_ID column are unique. Hence, this column would not add any value to our analysis. Let's drop this column.
```{r}
data <- data[ , !(names(data) %in% c("Booking_ID"))]
head(data)
```

## Check for missing values
```{r}
sum(is.na(data))
```

## Summary Statistics for Numerical Columns
```{r}
num_cols <- sapply(data, is.numeric)
summary(data[ , num_cols])
```
**Observations:**

- The average price per room is 103.4 euro. It has a wide range up to 540 euro. The median is 99 euro, so there seems to be high-price outlier rooms.

- There are entries where the average price per room is zero. This needs to be examined further. Perhaps promotional deals?

- The mean number of week nights for a stay is 2.2 while it is 0.81 weekend days.

- The average number of previous cancellations is 0.02. The quartiles are all zero and the maximum is 58, indicating possibly sparse data.

- The average lead time is 85 days. The median is 57 and the max is 443. There are high lead time outliers and the distribution is likely skewed to the right.

## Bar plot to show distribution of cancellations vs non-cancellations
```{r}
print(ggplot(data, aes(x = booking_status)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Booking Status Distribution (Canceled vs Not Canceled)",
       x = "Booking Status", y = "Count"))
```

## Box plot for Lead Time vs Cancellations:
```{r}
# Box plot for lead time vs. cancellations
print(ggplot(data, aes(x = booking_status, y = lead_time, fill = booking_status)) +
  geom_boxplot() +
  labs(title = "Lead Time by Booking Status", x = "Booking Status", y = "Lead Time"))
```

## Average Price per Room vs Cancellations
```{r}
# Histogram for average price per room by booking status
print(ggplot(data, aes(x = avg_price_per_room, fill = booking_status)) +
  geom_histogram(bins = 30, position = "dodge") +
  labs(title = "Distribution of Average Price per Room by Booking Status",
       x = "Average Price per Room", y = "Count"))
```



## Seasonality Analysis:
```{r}
# Define seasons based on arrival month
data$season <- case_when(
  data$arrival_month %in% c(3, 4, 5) ~ "Spring",
  data$arrival_month %in% c(6, 7, 8) ~ "Summer",
  data$arrival_month %in% c(9, 10, 11) ~ "Fall",
  data$arrival_month %in% c(12, 1, 2) ~ "Winter",
  TRUE ~ "Unknown"
)
# Convert booking_status to a binary variable
data$booking_status_binary <- ifelse(data$booking_status == "Canceled", 1, 0)
```

### Creating subsets for each season:
```{r}
# Create subsets for each season
spring_data <- subset(data, season == "Spring")
summer_data <- subset(data, season == "Summer")
fall_data <- subset(data, season == "Fall")
winter_data <- subset(data, season == "Winter")
```
